---
title: "St. Louis Crime Analysis"
author: "Jack Colpitt"
date: "11/10/2023"
output:
    rmdformats::readthedown:
      self_contained: true
      thumbnails: false
      lightbox: true
      gallery: true
      highlight: tango
      always_allow_html: yes
      code_folding: hide
---

```{css, echo=FALSE}


h1 {
  color: black !important;
}

h2 {
  color: black !important;
  <!-- background-color:#ffeec5; -->
  <!-- padding: 5px; -->
  <!-- border-radius: 10px; -->
  <!-- border: 1px solid black; -->
}

#sidebar {
  background-color: black;
}

#sidebar h2 {
  background-color:#ffeec5;
  border-radius: 0px;
  border: 0px;
}

#sidebar h2 a {
  color:black;
}

#toc {
  background-color:black;
}

#postamble {
  border-top: black;
}
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We will be performing exploratory data analysis for three types of recorded criminal activity in the St. Louis area. The study focuses on arson, DUIs, and homicides. The purpose of this study is to answer questions about that data and gain familiarity for continuing analysis. Our methods in the analysis will use descriptive statistical techniques to answer questions, such as, "where did these crimes take place?" (Figure 4), "On average, how often are they happening?" (Figure 2 & Figure 3), and "Which crime is happening the most?" (Figure 1).

## Required R Packages

The code cell below indicates the R packages used in the analysis.

```{r echo=TRUE, message = FALSE, class.source = 'fold-show'}
#install and load packages here
library(ggplot2)
library(maptools)
library(spatstat)
library(leaflet)
library(dplyr)
library(doBy)
library(rgdal)
library(ggmap)
library(gridExtra)
library(sp)
library(RColorBrewer)
library(DT)
```

## Methods

In our analysis we are leveraging exploratory data analysis and visualization with the R programming language to gain an understanding of our chosen data set and eventually suggest a hypothesis if deemed appropriate (Rogerson, 2011). The datatable shown in Table 1 allows us to preview the selected variables and observations. The table was filtered to show crime types, and frequency per month. Further insights are gained by showing trends in total number of crimes based on category (Figure 1) and crimes per month with an overlaid density curve in Figure 2 to quickly spot heightened observations. Figure 3 separates the data based on type, month, and year. A spatial display using R Leaflet is in Figure 4 where a possible hypothesis of crime type clustering could be present. Lastly, descriptive statistics are used for finding the mean, median, mode, standard deviation, and outliers for grasping a better understanding of our data. 

## DataTable
```{r, echo=TRUE}
# set working directories
file_dir_crime <- "crime/"

# set other directories
file_dir_gis <- "gis/"

#read in crime file
#concatenate the filedircrime (set earlier to your directory) with the filename using paste()
crime_file <- paste(file_dir_crime,"crimeStLouis20132014b.csv", sep = "")
crime_file_agg <- paste(file_dir_crime,"crimeStLouis20132014b_agg.csv", sep = "")

# read data into csv 
crimesstl <- read.csv(crime_file, header=TRUE,sep=",")
crimesstl_agg <- read.csv(crime_file_agg, header=TRUE,sep=",") 
 
#create a list of the unique crime types in the data set and view what these are so that you can select using these so that you can explore their distributions.
listcrimetype <-unique(crimesstl$crimetype)

#returns a datatable of the loaded csv
sub_crime_agg <- crimesstl_agg %>%
  select(count, crimetype, month)

datatable(sub_crime_agg, width = "auto")
```
**Table 1.** By utilizing the datatable we can order count rate in a descending order and see that DUIs in May have the highest hit rate for a single crime type based on monthly intervals. 

## What type of crimes occur more than others?

One might assume from observations in Table 1 that DUIs would be the highest counted crime in our study. By summing the count columns from all crimes in 2013 and 2014 we actually see that arson has the highest counted crime type in comparison to homicide an DUIs. The number for this summation and conclusion is visible in Figure 1. From this conclusion I would make an effort to understand what event took place to increase the level of arson above the number of DUIs.

```{r echo=TRUE}
#Barchart of crimes by crimetype
counts <- table(crimesstl$crimetype)
barplot(counts, col = "cornflowerblue", main = "Number of Crimes by Crime Type", xlab="Crime Type", ylab="Number of Crimes")
```
**Figure 1.** In combined years 2013 and 2014 Arson is the most common crime followed by DUI, and homicide. This chart is informative, and accurate, but may be misleading to conclude that arson is consistently more common than DUIs in St. Louis.

## Temporal patterns associated with each crime 

Now let's dive deeper into the data, and start to visualize information based on when it took place. Perhaps with understanding when the events took place, we can start to see if there is a particular reason or event that is skewing our data to a misguided interpretation. I would hypothesis in this stage of the process that an event occurred in 2013 or 2014 that led to an extreme level of arson.

```{r echo=TRUE, message=FALSE, warning=FALSE}
#Histograms of crimes by observations per month

ggplot(crimesstl, aes(x = month)) + 
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  geom_density(aes(y = after_stat(count)), color = "red", linewidth = 1) +
  labs(title = "Histogram of Crimes by Month with Density Curve",
       x = "Month",
       y = "Frequency") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_x_continuous(breaks = 2:12)
```
**Figure 2.** By combining all crime types in 2013 and 2014, most observations took place in the month of August.

Now that we have seen the population of our data summed into one chart that gives an interesting perspective that the month of august experienced an above average number of crimes. We know very little of what kind of events took place. Utilizing a facet wrapped plot based on crime type and year we can see when each category of crimes experienced peaks and troughs based on monthly intervals. 

```{r echo=TRUE}
crimess2013 <- subset(crimesstl, year == "2013")
crimess2014 <- subset(crimesstl, year == "2014")

counts_2013 <- table(crimess2013$month, crimess2013$crimetype)
counts_df_2013 <- as.data.frame(counts_2013)
colnames(counts_df_2013) <- c("Month", "CrimeType", "Count")

# Create the faceted barplot
plot_2013 <- ggplot(counts_df_2013, aes(x = Month, y = Count, fill = CrimeType)) +
              geom_bar(stat = "identity", show.legend = FALSE) +
              labs(title = "2013",
                   x = "Month",
                   y = "Number of Crimes") +
              scale_fill_brewer(palette = "Set1") +  # You can change the palette if needed
              facet_wrap(~ CrimeType, scales = "free_x", ncol = 1)

counts_2014 <- table(crimess2014$month, crimess2014$crimetype)
counts_df_2014 <- as.data.frame(counts_2014)
colnames(counts_df_2014) <- c("Month", "CrimeType", "Count")

# Create the faceted barplot
plot_2014 <- ggplot(counts_df_2014, aes(x = Month, y = Count, fill = CrimeType)) +
              geom_bar(stat = "identity") +
              labs(title = "2014",
                   x = "Month",
                   y = "Number of Crimes") +
              scale_fill_brewer(palette = "Set1") +  # You can change the palette if needed
              facet_wrap(~ CrimeType, scales = "free_x", ncol = 1)

grid.arrange(plot_2013, plot_2014, nrow=1,ncol=2, widths = c(.75, 1), top = "Crime Type Event Comparison")
```
**Figure 3.** Arson levels are heightened in August of 2013. The ggplot library and the facet_wrap() function was used to seperate the variables and visualize the information side by side. 

Arson events in August and September of 2013 are appearing at elevated levels.

## Descriptive Statistics



## Where did the crimes take place?

Diplaying point location based on attributes might allow us to create potential hypothesis if location is a determining factor for the results we have seen so far. Clusters of crime type can be seen across St. Louis in Figure 4 below. 

```{r echo=TRUE, message=FALSE, warning = FALSE}
#This will create a map with all of the points
gis_file <- paste(file_dir_gis,"stl_boundary_ll.shp", sep="")
#Read the St Louis Boundary Shapefile
StLouisBND <- readOGR(gis_file, layer = "stl_boundary_ll", GDAL1_integer64_policy = FALSE, verbose = FALSE)

crimess22 <- subset(crimesstl, crimetype == "arson")
crimess33 <- subset(crimesstl, crimetype == "dui")
crimess44 <- subset(crimesstl, crimetype == "homicide")
#Create an interactive map for each crimetype
#To view a particular crime you will want to create a map using a subset of the data.
#Or use leaflet and include a background map to put the crime in context.
 
leaflet(crimesstl) %>%
  addTiles() %>%
  addPolygons(data=StLouisBND, 
              color = "#444444", 
              weight = 3, 
              smoothFactor = 0.5,
              opacity = 1.0, 
              fillOpacity = 0.5, 
              fill= FALSE,
              highlightOptions = highlightOptions(color = "white", weight = 2,
                                                      bringToFront = TRUE),
              group='St. Louis') %>%
   
  addCircles(data = crimess22,
             color = "red",
             lng = crimess22$xL, 
             lat = crimess22$yL, 
             weight = 5, 
             radius = 10,
             group = "Arson",
    popup = paste0("Crime type: ", as.character(crimess22$crimetype), 
                              "; Month: ", as.character(crimess22$month)))  %>%
 
 
#Turn these on when you are ready to view them. To do so remove the # and switch off the addCircles above by adding a #
  addCircles(data = crimess33,
             color = "blue",
             lng = crimess33$xL, 
             lat = crimess33$yL, 
             weight = 5, 
             radius = 10,
             group = 'DUI',
      popup = paste0("Crime type: ", as.character(crimess33$crimetype), 
                                "; Month: ", as.character(crimess33$month)))  %>%
   
  addCircles(data = crimess44,
             color = "green",
             lng = crimess44$xL, 
             lat = crimess44$yL, 
             weight = 5, 
             radius = 10,
             group = 'Homicide',
      popup = paste0("Crime type: ", as.character(crimess44$crimetype), 
                                "; Month: ", as.character(crimess44$month)))  %>%

  addLayersControl(
          overlayGroups =c('St. Louis', 'Arson', 'DUI', 'Homicide'),
          options = layersControlOptions(collapsed=FALSE)
          ) %>%
  
  addLegend("bottomleft", 
           colors = c("#444444", "red", "blue", "green"), 
           labels = c("St. Louis", "Arson", "DUI", "Homicide"), 
           title = "Crime Type",
           opacity = 1
  )
```
**Figure 4.** Strong spatial dependence is portrayed in arson and homicides, while the opposite is viewed for DUIs. 

Spatial dependent patterns for arson and homicide might lead us to hypothesize an underlying reason for the clustering of events. DUIs are the other hand lack spatial dependence and the variation is more dispersed across the city. Therefore spatial autocorrelation is higher regarding arson and homicides, than DUIs. 

#### If theres time make a graduated color palette for arson to see the if the events were happening in waves across the city.

## Descriptive Statistics

Boxplots are informative diagrams that allow us to see the quartile ranges (Q1, Q2 - Median, and Q3), the low and high end the range, and outliers in the data. IQR is determined by Q3 - Q1, and values that exist below Q1 − 1.5(IQR) as well as values above Q3 + 1.5(IQR) are deemed as outliers. 

```{r echo=TRUE}
#Summarise data
summary(crimesstl)
 
#Make a data frame (a data structure) with crimes by crime type
dt <- data.frame(cnt=crimesstl$count, group=crimesstl$crimetype)
#save these grouped data to a variable so you can use it other commands
grp <- group_by(dt, group)
 
#Summarise data from library (dplyr)
#Summarise the number of counts for each group
summarise(grp, sum=sum(cnt))
#transpose the table
tapply(crimesstl$count, crimesstl$crimetype,sum)
```

```{r echo=TRUE}
#BoxPlots are useful for comparing data. 
#Use the dataset crimeStLouis20132014b_agg.csv.  
#These data are aggregated by neighbourhood.  
agg_crime_file <-paste(file_dir_crime,"crimeStLouis20132014b_agg.csv", sep = "")
#check everything worked ok with accessing the file
#file.exists(agg_crime_file)
crimesstlagg <- read.csv(agg_crime_file, header=TRUE,sep=",")
 
#Compare crimetypes
boxplot(count~crimetype, data=crimesstlagg,main="Boxplots According to Crime Type",
            xlab="Crime Type", ylab="Number of Crimes",
            col="cornsilk", border="brown", pch=19)
```
**Figure 5.** 
