---
title: "GEOG 586 Lab 3B"
author: "Jack Colpitt"
date: "10/22/2023"
output:
    html_notebook
---

# St. Louis Crime Point Pattern Analysis

#### Packages used in project
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(spatstat)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(gridExtra)
```

What were your overall findings about crime in your study area?

Where did crimes take place? Was there a particular spatial distribution associated with each of the crimes? Did they take place in a particular part of the city?

Did the different crimes you assessed cluster in different parts of the study area or did they occur in the same parts of the city? 

What additional types of analysis would be useful?

```{r echo=FALSE, warning=FALSE}
#crime data
file_dir_gis <- paste0(getwd(), "/Geog586_Les3_Project/gis/")
gis_file <- paste0(file_dir_gis, "stl20132014sube.shp", sep="")

#study area boundary
crimes_shp_prj<- read_sf(gis_file)
crimes_sf <- st_as_sf(crimes_shp_prj)

gis_file_2 <-paste0(file_dir_gis,"stl_boundary.shp", sep="")
StLouis_BND_prj <-read_sf(gis_file_2)
Sbnd <-st_as_sf(StLouis_BND_prj)

#Set up the analysis environment by creating an analysis window from the city boundary
Sbnd <-st_as_sf(StLouis_BND_prj)
#plot(Sbnd)

#add the crime events to the analysis window
crimesppp <-as.ppp(crimes_shp_prj)

#plot the points to the boundary as a solid red circle one-half the default plotting size
#points(crimesppp$x, crimesppp$y, pch = 19, cex=0.5, col = "blue")

geom_crimes_sf <- crimes_sf %>%
  mutate(lon = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2]) %>%
  select(crimet2, lon, lat) 

#all_crimes <- ggplot() +
#  geom_sf(data = Sbnd) +
#  geom_point(data = geom_crimes_sf, aes(x = lon, y = lat)) +
#  labs(title = "St. Louis Crimes",  color = "") +
#  theme(legend.position = "none",
#        panel.grid = element_blank(),
#        axis.title = element_blank(),
#        axis.text = element_blank(),
#        axis.ticks = element_blank(),
#        panel.background = element_blank()) 



sf_homicide <-geom_crimes_sf %>%
  filter(crimet2 == 'homicide')

homicides <- ggplot() +
  geom_sf(data = Sbnd) +
  geom_point(data = sf_homicide, aes(x = lon, y = lat)) +
  labs(title = "St. Louis Homicides", subtitle="2013 - 2014",  color = "") +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank()) 

homicides
```
**Figure 1.** Homicide crimes plotted against all crime types for St. Louis. 

## Kernel Density Analysis

Kernel estimation allows for an analysis of intensity at all locations in the study area, not just at the location of the events (Lloyd 2011). Making it a great choice for understand first order effects.
```{r, echo=FALSE}
par(mfrow=c(1, 3))
xhomicide <-crimesppp[crimes_shp_prj$crimet2=="homicide"]

K1 <-density(xhomicide) # Using the default bandwidth
plot(K1, main=NULL, las=1)
plot(xhomicide, pch=19, col="white", use.marks=FALSE, cex=0.75, add=TRUE)
contour(K1, add=TRUE, col = "white")
#add a label to the top of the plot
mtext("default (Oversmoothed)") 

#Density–changing bandwidth
K2 <-density(crimesppp, sigma=500) # set different bandwidths. This data is in projected feet.
plot(K2, main=NULL)
plot(xhomicide, pch=19, col="white", use.marks=FALSE, cex=0.75, add=TRUE)
contour(K2, add=TRUE, col = "white")
mtext("500 ft (Undersmoothed)") 

KDE_opt <-bw.diggle(xhomicide)
K3 <-density(xhomicide, sigma=KDE_opt) # Using the diggle bandwidth
plot(K3, main=NULL)
plot(xhomicide, pch=19, col="white", use.marks=FALSE, cex=0.75, add=TRUE)
contour(K3, add=TRUE, col = "white")

diggle_bandwidth <-toString(KDE_opt, width = 5)
mtext("Suggested Optimal")
```
**Figure 2.** Range of variation in bandwidth estimation for the kernel density estimation. 

# Nearest Neighbour Analysis
```{r}

nnd <-nndist.ppp(xhomicide)
hist(nnd)
summary(nnd)
 
mnnd <-mean(nnd)
exp_nnd <-(0.5/ sqrt(xhomicide$n / area.owin(Sbnd)))
mnnd / exp_nnd
```

# Quadrat Analysis

```{r}
#Quadrat Analysis 
q <-quadratcount(xhomicide, 4, 8) 
plot(q) 
 
#Add intensity of each quadrat
#Note, depending on your version of R, this line of code may not work correctly 
plot(intensity(q, image=TRUE), main=NULL, las=1) 
 
#perform the significance test 
quadrat.test(xhomicide, 4, 8)
```

# Distance Based Analysis with Monte Carlo Assessment

```{r}

```