---
title: "GEOG 586 Lab 3B"
author: "Jack Colpitt"
date: "10/22/2023"
output:
    html_notebook
---

# St. Louis Crime Point Pattern Analysis

#### Packages used in project
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(spatstat)
library(tidyverse)
```

What were your overall findings about crime in your study area?

Where did crimes take place? Was there a particular spatial distribution associated with each of the crimes? Did they take place in a particular part of the city?

Did the different crimes you assessed cluster in different parts of the study area or did they occur in the same parts of the city? 

What additional types of analysis would be useful?

```{r echo=FALSE, warning=FALSE}
#crime data
file_dir_gis <- paste0(getwd(), "/Geog586_Les3_Project/gis/")
gis_file <- paste0(file_dir_gis, "stl20132014sube.shp", sep="")

#study area boundary
crimes_shp_prj<- read_sf(gis_file)
gis_file_2 <-paste0(file_dir_gis,"stl_boundary.shp", sep="")
StLouis_BND_prj <-read_sf(gis_file_2)

#Set up the analysis environment by creating an analysis window from the city boundary
Sbnd <-as.owin(StLouis_BND_prj)
#check the shapefile will plot correctly
plot(Sbnd)

#add the crime events to the analysis window
crimesppp <-as.ppp(crimes_shp_prj)

#plot the points to the boundary as a solid red circle one-half the default plotting size
points(crimesppp$x, crimesppp$y, pch = 19, cex=0.5, col = "red")
```
## Kernel Density Analysis
```{r}
#KDE Analysis
 
#Density
K1 <-density(crimesppp) # Using the default bandwidth
plot(K1, main=NULL, las=1)
contour(K1, add=TRUE, col = "white")
#add a label to the top of the plot
mtext("default bandwidth")
```

```{r}
#Density–changing bandwidth
K2 <-density(crimesppp, sigma=500) # set different bandwidths. This data is in projected feet.
plot(K2, main=NULL)
contour(K2, add=TRUE, col = "white")
mtext("500 feet") 
```
```{r}
#Density–optimal bandwidth
 
#R provides a function that will suggest an optimal bandwidth to use:
#You can replicate this code to find out the other bandwidth values. when you examine 
#individual crime types later in the lesson.
#KDE_opt <-bw.diggle(crimesppp)
#KDE_opt

K3 <-density(crimesppp, sigma=KDE_opt) # Using the diggle bandwidth
plot(K3, main=NULL)
contour(K3, add=TRUE, col = "white")

# Convert the bandwidth vector to a character string with 5 character places 
diggle_bandwidth <-toString(KDE_opt, width = 5) 
# Add the bandwidth value to the plot title 
mtext(diggle_bandwidth) 

#Now that you are familiar with how KDE works, lets create a KDE for one particular crime type. 
 
#To select an individual crime specify the crimetype as follows:
xhomicide <-crimesppp[crimes_shp_prj$crimet2=="homicide"]

KHom<-density(xhomicide) # Using the default bandwidth
plot(KHom, main=NULL, las=1)
contour(KHom, add=TRUE, col = "white")
plot(xhomicide, pch=19, col="white", use.marks=FALSE, cex=0.75, add=TRUE) #the pch=19 sets the symbol type 
#to a solid filled circle, the col="white" sets the fill color for the symbols,
#we set use.marks=FALSE so that we aren't plotting a unique symbol (ie square, circle, etc) for each observation, 
#cex=0.75 sets the size of the symbol to be at 75% of the default size, 
#and add=TRUE draws the crime points on top of the existing density surface.
mtext("homicide default bandwidth") 
```